Theory of phonon polaritons in polar dielectrics

### 2018-08-03 paper.tex
commit

There are nevertheless some outstanding questions regarding the behavior of phonon polaritons. Here, we present a few. First, can the very successful Lorentz oscillator model ever break down? One particular mechanism of breakdown, which could be present in the case of extremely confined polaritons, concerns effects coming from spatial dispersion. Such short phonon polariton wavelengths are indeed predicted by classical electrodynamics of a polar dielectric, especially in the case of a very thin film (nm-thick). Second, what are the properties of phonon polaritons in reduced dimensions, such as 2D polar insulators? Can they support extremely confined polaritonic modes, in analogy to extremely confined plasmonic modes in 2D plasmonic materials such as graphene? In general, are the properties of optical phonons inherited by the phonon polariton? Finally, what is the influence of non-trivial spatial shaping of phonons on the resulting polariton? Examples of non-trivial spatial shaping include chiral phonons, interface phonons, and mechanical heterostructurs. The ability to design new phonon polariton materials by designing the optical phonon band structure allows a predictive way to generate polaritonic nano-materials.

The way to proceed on all of these questions is an ab-initio framework to predict the phonon contribution to the dielectric tensor in arbitrary settings. Frameworks exist for calculating the local permittivity coming from phonons in a 3D bulk, as in \cite{Gonze1997dynamical}, and implemented in software packages such as ABINIT \cite{abinit1,abinit2,abinit3}. Here, we develop a unified 
theoretical framework to analyze phonon-polaritons in materials, which can address effects of spatial dispersion as well as effects coming from arbitrary dimensionality. Specifically, we develop a means to arrive at the dielectric function associated with optical phonons in two- and three-dimensional materials. We pay close attention to develop a description of the optical response that shows clearly the parallels to the description of plasmonic response from first principles. From this theory, we not only accurately predict surface phonon polaritons in thick materials where a bulk permittivity is valid  but also provide tools to predict phonon polariton dispersion in monolayers and anticipate when spatially nonlocal corrections to the phonon contribution to the dielectric function become relevant.

@@

We start by developing the theory of phonon polaritons in a 3D bulk in a way that is parallel to the linear response theory of electronic contributions to the permittivity and plasmonic properties. Here, the matter subsystem which couples to light is the ionic lattice of the crystal. Thus, we develop the linear response theory of an ionic lattice to an electromagnetic field. The interaction Hamiltonian of the lattice with the electromagnetic field is taken as
\begin{equation}
H_{int} = -\int d^3x ~\mathbf{E}(\mathbf{x},t)\cdot\mathbf{P}(\mathbf{x},t),
\end{equation}
where $\mathbf{E}$ is a driving electric field, and $\mathbf{P}$ is the polarization associated with the lattice. We write the polarization as a sum of contributions from all unit cells in the system. Namely
\begin{equation}
\mathbf{P}(\mathbf{r},t) = \sum\limits_{\text{sites, }\mathbf{R}}\mathbf{d}(\mathbf{R},t)\delta(\mathbf{r}-\mathbf{R}),
\end{equation}
where $\mathbf{d}(\mathbf{R},t)$ is the time-dependent dipole moment associated with lattice site $\mathbf{R}$. In expressing the polarization in this fashion, we are assuming that the displacement of the ions is far smaller than the size of the unit cell, which is, even in a deeply anharmonic regime, a good approximation. This said, we operate in the harmonic regime.  The dipole moment is given by $\mathbf{d}(\mathbf{R},t) = \sum\limits_{\text{basis sites,}\kappa} \mathbf{b}_{\kappa} \mathbf{Z}_{\kappa}\mathbf{u}(\mathbf{R}_{\kappa} \equiv \mathbf{R}+\mathbf{b}_{\kappa},t)$, where $\mathbf{Z}_{\kappa}$ is the tensor of Born effective charges for atom $\kappa$ in the unit cell at basis vector $\mathbf{b}_{\kappa}$ and $\mathbf{u}(\mathbf{R}_{\kappa},t)$ is the displacement of atom $\kappa$ at basis site $\mathbf{R}_{\kappa}$. Note that the Born effective charges are both tensorial and site-dependent.  This displacement can be expanded in phonon modes as
\begin{equation}
\mathbf{u}(\mathbf{R}_{\kappa},t) = \sum\limits_{\mathbf{q}\sigma}\sqrt{\frac{\hbar}{2M_{\kappa}N\omega_{\mathbf{q}\sigma}}}\left( e^{i\mathbf{q}\cdot\mathbf{R}_{\kappa}-i\omega_{\mathbf{q}\sigma}t}\hat{e}_{\mathbf{q}\sigma}(\mathbf{b}_{\kappa})a_{\mathbf{q}\sigma} + \text{h.c.} \right),
\end{equation}
where $M_{\kappa}$ is the mass of atom $\kappa$, $\omega_{qs}$ is the phonon frequency at wavevector $\mathbf{q}$ and branch $s$, and $N$ is the number of unit cells.
Assuming that the field is weak, the expectation value of the polarization in the lattice will depend linearly on the total field acting on the ions: this field is the sum of the external field with the field resulting from induced electronic and ionic polarization, and it is this field which appears in the interaction Hamiltonian of Equation (1). We define the linear-response relation for a bulk crystal in Fourier space as $\langle\mathbf{P}(\mathbf{q},\omega)\rangle = \epsilon_0\boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}(\mathbf{q},\omega)$, where we refer to $\boldsymbol{\Pi}(\mathbf{q},\omega)$ as the polarization-polarization response function.

@@

Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on "nonlocality" in the phonon contribution to the dielectric function. Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion.  Therefore, it is possible that the phonon polariton resonances red-shift or blue-shift, depending on how the relative position of the phonons at finite-wavevector versus zero wavevector. For example, in 4H-SiC, upon moving along the $\Gamma-L$ direction of the Brillouin zone, the optical phonon which determines the phonon polariton, calculated to be at 764 cm$^{-1}$ at $\Gamma$, red-shifts by 2  cm$^{-1}$ at 1/10 of the way along the $\Gamma-L$ direction, by 5 cm$^{-1}$ at 1/10 of the way along the $\Gamma-L$ direction, and 13 cm$^{-1}$ at 1/2 of the way along the $\Gamma-L$ direction. This will lead to a red-shift of the Reststrahlen bands by these amounts.

@@

@@

\section{Linear response of phonon polaritons in two dimensions}
We now move to generalize the derivation of the previous section to describe the optical phonon contribution to Maxwell's equations coming from a two-dimensional material. Unlike the previous section, we will not demonstrate the results of the calculation on a particular material. Here, we are only interested in finding an expression for the optical response of the surface in terms of ab-initio-derivable parameters, as well as the general form of the dispersion relation of phonon-polaritons in 2D.  Because of the change of symmetry in the system, we will not be able to perform the analysis by Fourier transforming Maxwell's equations in three dimensions. However, we will apply a technique which is often used to analyze electromagnetic modes in 2D systems such as plasmons in 2D conductors. An example of the technique is shown in detail in \cite{jablan2009plasmonics}, but we summarize the essential details here.

@@

To connect this relation to the discussion of the three-dimensional case, we write a microscopic form for the polarization response function of the ionic lattice. It is simply
\begin{equation}
\boldsymbol{\Pi}(\mathbf{q},\omega) =  \frac{1}{\epsilon_0 A}\sum\limits_{m,n}\frac{\mathbf{P}_{mn}(\mathbf{q})\otimes\mathbf{P}_{nm}(\mathbf{q})}{\hbar\omega + E_{nm}+i0^+}\left(e^{\beta E_m}-e^{\beta E_n} \right),
\end{equation}
where the only change is to replace a normalization volume with a normalization area $A$. Here, $\mathbf{P}(\mathbf{q})$ is the 2D Fourier transform of the polarization density. As the form of the displacement of the ions is still given by Equation (3), $\mathbf{P}(\mathbf{q})$ is still given by Equations (7-10), but with the Born charges and eigendisplacements appropriate to the 2D system of interest. Note that although the lattice is two dimensional, in general, the ions may be displaced in any three directions and thus $\boldsymbol{\Pi}$ is still a $3\times3$ matrix. Following the same kind of reasoning that lead to Equation (13), if we parameterize the effect of the different masses and Born charges in the unit cell by $M^*$ and $Q^*$, we will find a form for the polarization response function which is given by the second term of Equation (13) but with $V$ replaced by $A$. Defining $N/A \equiv n_s$, Equation (14) for the dispersion relation of 2D phonon polaritons tells us that
\begin{equation}
\omega = \sqrt{\omega_{TO}^2+\frac{n_sQ^{*2}}{2M^*\epsilon_0\epsilon}q}.
\end{equation}
Note that this dispersion has a strong resemblance to the plasmon dispersion relation of two-dimensional electron gases, which is normally of the form\cite{stern1967polarizability} $\omega = \sqrt{\frac{n_s e^2  }{2m\epsilon_0\epsilon}q}$ . The only difference is the replacement of electron parameters (density, charge, mass) by the corresponding lattice parameters in the phononic case, and a shift in the minimum frequency by the transverse optical phonon frequency.

We note that the dispersion relation of Equation (8) can in fact be obtained by taking the infinitely thin-limit of a bulk polar dielectric, and finding the dispersion of the resulting phonon polaritons in the quasi-electrostatic limit. In particular, for an isotropic polar dielectric of thickness $2d$ with boundaries at $z=\pm d$, solving Laplace's equations yields as the dispersion of the even parity mode:
\begin{equation}
\tanh(qd) = -\frac{\epsilon}{\epsilon(\omega)} \rightarrow qd\epsilon(\omega) = -\epsilon,
\end{equation}
with $\epsilon$ the dielectric constant of the surrounding background and $\epsilon(\omega)$ that of the polar dielectric. The right-hand side is obtained in the limit of $d\rightarrow 0$. Taking Equation (13) for the dielectric function in bulk, we have that (in the absence of losses) $\epsilon + qd\epsilon_{\infty} + q\frac{NQ^{*2}}{2\epsilon_0 M^*A}\frac{1}{\omega_{TO}^2-\omega^2} = 0$, where we have identified $V = 2Ad$. In the limit of small $d$, i.e., $qd \ll 1$, we recover Equations (16) and (18) with $\Pi = \frac{NQ^{*2}}{\epsilon_0 MA}\frac{1}{\omega^2_{TO}-\omega^2}$. A useful result which follows is that we may also derive another important figure of merit, the $p-$polarized reflectivity of a 2D polar dielectric, which is related to the local density of states of the polar dielectric, which quantifies the strength of coupling to external probes such as a nano-antenna or a quantum emitter. By considering the infinitely thin limit of a polar dielectric with dielectric function $\epsilon(\omega)$ and defining $\epsilon(\omega) = \frac{\Pi}{2d}$, we immediately find that the p-polarized reflectivity is given by
\begin{equation}
r_p = \frac{1}{1+\frac{2\epsilon}{q\Pi}}.
\end{equation}

>>>>

There are nevertheless some outstanding questions regarding the behavior of phonon polaritons. Here, we present a few. First, can the very successful Lorentz oscillator model ever break down? One particular mechanism of breakdown, which could be present in the case of extremely confined polaritons, concerns effects coming from spatial dispersion. Such short phonon polariton wavelengths are indeed predicted by classical electrodynamics of a polar dielectric, especially in the case of a very thin film (nm-thick). Second, what are the properties of phonon polaritons in reduced dimensions, such as 2D polar insulators? Can they support extremely confined polaritonic modes, in analogy to extremely confined plasmonic modes in 2D plasmonic materials such as graphene?  Finally, what is the influence of non-trivial spatial shaping of phonons on the resulting polariton? Examples of non-trivial spatial shaping include chiral phonons, interface phonons, and mechanical heterostructurs. The ability to design new phonon polariton materials by designing the optical phonon band structure allows a predictive way to generate polaritonic nano-materials.

The way to proceed on all of these questions is an ab-initio framework to predict the phonon contribution to the dielectric tensor in arbitrary settings. Frameworks exist for calculating the local permittivity coming from phonons in a 3D bulk, as in \cite{Gonze1997dynamical}, and implemented in software packages such as ABINIT \cite{abinit1,abinit2,abinit3}. Here, we develop  a unified 
theoretical framework based on linear response to analyze phonon-polaritons in materials, which can address effects of spatial dispersion as well as effects coming from arbitrary dimensionality. Specifically, we develop a means to arrive at the dielectric function associated with optical phonons in two- and three-dimensional materials. We pay close attention to develop a description of the optical response that shows clearly the parallels to the description of plasmonic response from first principles, and allows us to point to the distinct kinds of non-localities that appear in phonons.  Using this theory, we first demonstrate that we may accurately predict the permittivity of bulk polar dielectrics such as silicon carbide. Using this framework, we show how to extend it to predict the properties of phonon polaritons in monolayers. Although we do not provide an ab-initio example of the phonon polariton dispersion in 2D, we, using reasonable simplifying assumptions, predict a universal form for the dispersion relation of phonon polaritons in two dimensions.

@@

We start by developing the theory of phonon polaritons in a 3D bulk in a way that is parallel to the linear response theory of electronic contributions to the permittivity and plasmonic properties. Here, the matter subsystem which couples to light is the ionic lattice of the crystal. Thus, we develop the linear response theory of an ionic lattice to an electromagnetic field. The interaction Hamiltonian of the lattice with the electromagnetic field is taken as
\begin{equation}
H_{int} = -\int d^3x ~\mathbf{E}(\mathbf{x},t)\cdot\mathbf{P}(\mathbf{x},t),
\end{equation}
where $\mathbf{E}$ is a driving electric field, and $\mathbf{P}$ is the polarization associated with the lattice. We write the polarization as a sum of contributions from all unit cells in the system. Namely
\begin{equation}
\mathbf{P}(\mathbf{r},t) = \sum\limits_{\text{sites, }\mathbf{R}}\mathbf{d}(\mathbf{R},t)\delta(\mathbf{r}-\mathbf{R}),
\end{equation}
where $\mathbf{d}(\mathbf{R},t)$ is the time-dependent dipole moment associated with lattice site $\mathbf{R}$. In expressing the polarization in this fashion, we are assuming that the displacement of the ions is far smaller than the size of the unit cell, which is, even in a deeply anharmonic regime, a good approximation. This said, we operate in the harmonic regime.  The dipole moment is given by $\mathbf{d}(\mathbf{R},t) = \sum\limits_{\text{basis sites,}\kappa}  \mathbf{Z}_{\kappa}\mathbf{u}(\mathbf{R}_{\kappa} \equiv \mathbf{R}+\mathbf{b}_{\kappa},t)$, where $\mathbf{Z}_{\kappa}$ is the tensor of Born effective charges for atom $\kappa$ in the unit cell at basis vector $\mathbf{b}_{\kappa}$ and $\mathbf{u}(\mathbf{R}_{\kappa},t)$ is the displacement of atom $\kappa$ at basis site $\mathbf{R}_{\kappa}$. Note that the Born effective charges are both tensorial and site-dependent.  This displacement can be expanded in phonon modes as
\begin{equation}
\mathbf{u}(\mathbf{R}_{\kappa},t) = \sum\limits_{\mathbf{q}\sigma}\sqrt{\frac{\hbar}{2M_{\kappa}N\omega_{\mathbf{q}\sigma}}}\left( e^{i\mathbf{q}\cdot\mathbf{R}_{\kappa}-i\omega_{\mathbf{q}\sigma}t}\hat{e}_{\mathbf{q}\sigma}(\mathbf{b}_{\kappa})a_{\mathbf{q}\sigma} + \text{h.c.} \right),
\end{equation}
where $M_{\kappa}$ is the mass of atom $\kappa$, $\omega_{qs}$ is the phonon frequency at wavevector $\mathbf{q}$ and branch $s$, and $N$ is the number of unit cells.
Assuming that the field is weak, the expectation value of the polarization in the lattice will depend linearly on the total field acting on the ions: this field is the sum of the external field with the field resulting from induced electronic and ionic polarization, and it is this field which appears in the interaction Hamiltonian of Equation (1). We define the linear-response relation for a bulk crystal in Fourier space as $\langle\mathbf{P}(\mathbf{q},\omega)\rangle = \epsilon_0\boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}(\mathbf{q},\omega)$, where we refer to $\boldsymbol{\Pi}(\mathbf{q},\omega)$ as the polarization-polarization response function.

@@

We now briefly comment on spatial nonlocality in the phonon contribution to the dielectric function. Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion.  Therefore, it is possible that the phonon polariton resonances red-shift or blue-shift, depending on how the relative position of the phonons at finite-wavevector versus zero wavevector. For example, in 4H-SiC, upon moving along the $\Gamma-L$ direction of the Brillouin zone, the optical phonon which determines the phonon polariton, calculated to be at 764 cm$^{-1}$ at $\Gamma$, red-shifts by 2  cm$^{-1}$ at 1/10 of the way along the $\Gamma-L$ direction, by 5 cm$^{-1}$ at 1/10 of the way along the $\Gamma-L$ direction, and 13 cm$^{-1}$ at 1/2 of the way along the $\Gamma-L$ direction. This will lead to a red-shift of the lower edge of the Reststrahlen band by these amounts. Beyond the shifts in the frequencies, the oscillator strengths ($\mathbf{S}_{\mathbf{q}\sigma}\otimes\mathbf{S}_{\mathbf{q}\sigma}$) of the phonon modes will also change as a function of frequency, and this will affect the strength of the phonon contribution to the permittivity.

@@

%Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on "nonlocality" in the phonon contribution to the dielectric function. Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion. As a result, the phonon contribution to the dielectric function sees contributions from the entire phonon band structure, including the large wavevector components. However, $\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$. In the vast majority of experiments that have been performed on phonon polaritons, the wavevector used to probe the polariton is on the order of $\frac{\omega}{c}$ which is much less than the wavevector scale at which the optical phonon dispersion varies, which is of the order of $\frac{\pi}{a}$ where $a$ is a lattice constant of the polar dielectric.  Even in scattering near-field optical microscope experiments, where one uses an AFM tip of radius of about $20$ nm to probe the polaritons, the wavevectors accessed would still be extremely small compared to the extent of the Brillouin zone. It would then seem almost hopeless to see nonlocal behavior for the phonon-polaritons. However, a recent experiment \cite{iranzo2018probing} has shown that plasmons in graphene, by means of a gold mirror, can be confined to dimensions of about 1 nm in the dimension transverse to the graphene sheet. Additionally, another recent experiment \cite{benz2016single} has leveraged so-called "picocavities", sub-nanometer gaps between a metal nanoparticle and metal film, to access very strong light-matter couplings and extreme variations in fields. These two experiments show a way to get large optically-accessed wavevectors to probe nonlocal behavior in polar dielectrics.
%>>>>>>> 3608b47a3ff25448fce9243424e9cbe4bd5888a3

@@

\section{Linear response of phonon polaritons in two dimensions}
We now move to generalize the derivation of the previous section to describe the optical phonon contribution to Maxwell's equations coming from a two-dimensional material. Unlike the previous section, we will not demonstrate the formalism on a particular material. Here, we are only interested in finding an expression for the optical response of the surface in terms of ab-initio-derivable parameters, as well as the general form of the dispersion relation of phonon-polaritons in 2D.  Because of the change of symmetry in the system, we will not be able to perform the analysis by Fourier transforming Maxwell's equations in three dimensions. However, we will apply a technique which is often used to analyze electromagnetic modes in 2D systems such as plasmons in 2D conductors. An example of the technique is shown in detail in \cite{jablan2009plasmonics}, but we summarize the essential details here.

@@

To connect this relation to the discussion of the three-dimensional case, we write a microscopic form for the polarization response function of the ionic lattice. It is simply
\begin{equation}
\boldsymbol{\Pi}(\mathbf{q},\omega) =  \frac{1}{\epsilon_0 A}\sum\limits_{m,n}\frac{\mathbf{P}_{mn}(\mathbf{q})\otimes\mathbf{P}_{nm}(\mathbf{q})}{\hbar\omega + E_{nm}+i0^+}\left(e^{\beta E_m}-e^{\beta E_n} \right),
\end{equation}
where the only change is to replace a normalization volume with a normalization area $A$. Here, $\mathbf{P}(\mathbf{q})$ is the 2D Fourier transform of the polarization density. As the form of the displacement of the ions is still given by Equation (3), $\mathbf{P}(\mathbf{q})$ is still given by Equations (7-10), but with the Born charges and eigendisplacements appropriate to the 2D system of interest. Note that although the lattice is two dimensional, in general, the ions may be displaced in any three directions and thus $\boldsymbol{\Pi}$ is still a $3\times3$ matrix.

Following the same kind of reasoning that lead to Equation (13), if we parameterize the effect of the different masses and Born charges in the unit cell by $M^*$ and $Q^*$, we will find a form for the polarization response function which is given by the second term of Equation (13) but with $V$ replaced by $A$. Defining $N/A \equiv n_s$, Equation (14) for the dispersion relation of 2D phonon polaritons tells us that
\begin{equation}
\omega = \sqrt{\omega_{TO}^2+\frac{n_sQ^{*2}}{2M^*\epsilon_0\epsilon}q}.
\end{equation}
Note that this dispersion has a strong resemblance to the plasmon dispersion relation of two-dimensional electron gases, which is normally of the form\cite{stern1967polarizability} $\omega = \sqrt{\frac{n_s e^2  }{2m\epsilon_0\epsilon}q}$ . The only difference is the replacement of electron parameters (density, charge, mass) by the corresponding lattice parameters in the phononic case, and a shift in the minimum frequency by the transverse optical phonon frequency. We note that in principle, there may be an additional background dielectric screening from the 2D sheet itself $\epsilon_{\infty}$, in which case, it would be added to $\boldsymbol{\Pi}$, leading to a shift in the dispersion with the overall form remaining the same.

We note that the dispersion relation of Equation (8) can in fact be obtained by taking the infinitely thin-limit of a bulk polar dielectric, and finding the dispersion of the resulting phonon polaritons in the quasi-electrostatic limit. In particular, for an isotropic polar dielectric of thickness $2d$ with boundaries at $z=\pm d$, solving Laplace's equations yields as the dispersion of the even parity mode:
\begin{equation}
\tanh(qd) = -\frac{\epsilon}{\epsilon(\omega)} \rightarrow qd\epsilon(\omega) = -\epsilon,
\end{equation}
with $\epsilon$ the dielectric constant of the surrounding background and $\epsilon(\omega)$ that of the polar dielectric. The right-hand side is obtained in the limit of $d\rightarrow 0$. Taking Equation (13) for the dielectric function in bulk, we have that (in the absence of losses) $\epsilon + qd\epsilon_{\infty} + q\frac{NQ^{*2}}{2\epsilon_0 M^*A}\frac{1}{\omega_{TO}^2-\omega^2} = 0$, where we have identified $V = 2Ad$. In the limit of small $d$, i.e., $qd \ll 1$, we recover Equations (16) and (18) with $\Pi = \frac{NQ^{*2}}{\epsilon_0 MA}\frac{1}{\omega^2_{TO}-\omega^2}$. A useful result which follows is that we may also derive another important figure of merit, the $p-$polarized reflectivity of a 2D polar dielectric, which is related to the local density of states of the polar dielectric, which quantifies the strength of coupling to external probes such as a nano-antenna or a quantum emitter. By considering the infinitely thin limit of a polar dielectric with dielectric function $\epsilon(\omega)$ and defining $\epsilon(\omega) = \frac{\Pi}{2d}$, we immediately find that the p-polarized reflectivity is given by
\begin{equation}
r_p = \frac{1}{1+\frac{2\epsilon}{q\Pi}},
\end{equation}
whose poles as expected correspond to the dispersion relation of Equation (16).
